---
title: "Landesque Capital and Ideal Distribution Models: Mapping Agricultural Suitability in the Sierra Nevada de Santa Marta, Colombia"
author: "Luis Miguel Soto Rodríguez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: show
    fig_width: 10
    fig_height: 8
  word_document:
    toc: true
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center", dpi = 300)
```

# 1. Configuration

```{r packages}
rm(list = ls()); gc()

required_packages <- c(
  "sf","terra","dplyr","ggplot2","pROC","broom","patchwork","spdep",
  "gstat","blockCV","viridis","scales","tidyr","gridExtra","forcats",
  "RColorBrewer","car","mgcv","boot","knitr","classInt","ggspatial",
  "ggpubr","rstatix","httr","jsonlite"
)

new_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if (length(new_packages) > 0) install.packages(new_packages, dependencies = TRUE)
suppressPackageStartupMessages({ lapply(required_packages, library, character.only = TRUE) })
```

```{r global-theme}
theme_jaa <- theme_bw(base_size = 12) + theme(
  text = element_text(color = "black"),
  axis.text = element_text(color = "black"),
  axis.title = element_text(color = "black", face = "bold"),
  plot.title = element_text(face = "bold", size = 13, color = "black"),
  plot.subtitle = element_text(size = 10, color = "black"),
  panel.grid.minor = element_blank(),
  legend.text = element_text(color = "black"),
  legend.title = element_text(color = "black", face = "bold"),
  strip.text = element_text(face = "bold", color = "black")
)
theme_set(theme_jaa)
```

# 2. Data loading from GitHub repository

```{r github-download}
api_url <- "https://api.github.com/repos/luismiguelsoto/Landesque_Capital_and_Ideal_Distribution_Models_in_Tairona_chiefdoms/contents/GIS"
respuesta <- GET(api_url)

if (status_code(respuesta) != 200) {
  stop("Connection to GitHub failed. Check the URL or API rate limits.")
}

archivos_github <- fromJSON(content(respuesta, "text", encoding = "UTF-8"))

folder <- file.path(tempdir(), "GIS_JAA")
dir.create(folder, showWarnings = FALSE, recursive = TRUE)

for (i in 1:nrow(archivos_github)) {
  if (!is.na(archivos_github$download_url[i])) {
    dest_file <- file.path(folder, archivos_github$name[i])
    if (!file.exists(dest_file)) {
      download.file(archivos_github$download_url[i], destfile = dest_file, mode = "wb", quiet = TRUE)
    }
  }
}

cat(">>> Files downloaded to:", folder, "\n")
cat(">>> Total files:", nrow(archivos_github), "\n")
```

```{r paths-and-data}
dem     <- terra::rast(file.path(folder, "DEM_POLYGON_52KM.tif"))
riofrio <- sf::st_read(file.path(folder, "RIO_FRIO_POLYGON_SURVEY.shp"), quiet = TRUE)
rivers  <- sf::st_read(file.path(folder, "RIVERS_POLYGON_52KM.shp"), quiet = TRUE)

sites_neg <- sf::st_read(file.path(folder, "NEGUANJE_POLYGONS_SITES.shp"), quiet = TRUE)
sites_bur <- sf::st_read(file.path(folder, "BURITACA_POLYGONS_SITES.shp"), quiet = TRUE)
sites_tai <- sf::st_read(file.path(folder, "TAIRONA_POLYGONS_SITES.shp"), quiet = TRUE)

terr_neg <- sf::st_read(file.path(folder, "NEGUANJE_STRUCTURES.shp"), quiet = TRUE)
terr_bur <- sf::st_read(file.path(folder, "BURITACA_STRUCTURES.shp"), quiet = TRUE)
terr_tai <- sf::st_read(file.path(folder, "TAIRONA_STRUCTURES.shp"), quiet = TRUE)

riofrio  <- sf::st_transform(riofrio, crs = sf::st_crs(dem))
rivers   <- sf::st_transform(rivers,  crs = sf::st_crs(dem))
sites_neg <- sf::st_transform(sites_neg, crs = sf::st_crs(dem))
sites_bur <- sf::st_transform(sites_bur, crs = sf::st_crs(dem))
sites_tai <- sf::st_transform(sites_tai, crs = sf::st_crs(dem))
terr_neg  <- sf::st_transform(terr_neg,  crs = sf::st_crs(dem))
terr_bur  <- sf::st_transform(terr_bur,  crs = sf::st_crs(dem))
terr_tai  <- sf::st_transform(terr_tai,  crs = sf::st_crs(dem))

sites_neg$Period <- "Neguanje (AD 100-700)"
sites_bur$Period <- "Buritaca (AD 700-1000)"
sites_tai$Period <- "Tairona (AD 1000-1600)"
sites_sf <- rbind(sites_neg, sites_bur, sites_tai)

terr_neg$Period <- "Neguanje (AD 100-700)"
terr_bur$Period <- "Buritaca (AD 700-1000)"
terr_tai$Period <- "Tairona (AD 1000-1600)"

common_cols <- c("UTM_East", "UTM_North", "Diam_Terra", "Period")
terrazas <- rbind(terr_neg[, common_cols], terr_bur[, common_cols], terr_tai[, common_cols])

all_structures_path <- file.path(folder, "ALL_STRUCTURES.shp")
if (file.exists(all_structures_path)) {
  all_structures <- sf::st_read(all_structures_path, quiet = TRUE)
  if (!is.na(sf::st_crs(all_structures)) && !is.na(sf::st_crs(dem)) &&
      sf::st_crs(all_structures) != sf::st_crs(dem)) {
    all_structures <- sf::st_transform(all_structures, sf::st_crs(dem))
  }
} else {
  all_structures <- terrazas
}

cat(">>> DATA LOADING SUMMARY:\n")
cat(" - DEM resolution:", round(terra::res(dem)[1], 1), "m\n")
cat(" - Survey area:", round(as.numeric(sf::st_area(riofrio)) / 1e6, 1), "km2\n")
cat(" - Total terraces:", nrow(terrazas), "\n")
cat("   Neguanje:", nrow(terr_neg), "| Buritaca:", nrow(terr_bur), "| Tairona:", nrow(terr_tai), "\n")
cat(" - Total sites:", nrow(sites_sf), "\n")
cat("   Neguanje:", nrow(sites_neg), "| Buritaca:", nrow(sites_bur), "| Tairona:", nrow(sites_tai), "\n")
cat(" - River segments:", nrow(rivers), "\n")
```

# 3. Helper functions

```{r helpers}
make_coord_key <- function(x, y, digits = 1) paste0(round(x, digits), "_", round(y, digits))

period_short <- function(x) dplyr::case_when(
  grepl("Neguanje", x) ~ "Neguanje\n(AD 100-700)",
  grepl("Buritaca", x) ~ "Buritaca\n(AD 700-1000)",
  grepl("Tairona",  x) ~ "Tairona\n(AD 1000-1600)",
  TRUE ~ x
)

period_factor <- function(x) {
  factor(x, levels = c(
    "Neguanje\n(AD 100-700)",
    "Buritaca\n(AD 700-1000)",
    "Tairona\n(AD 1000-1600)"
  ))
}

make_glm_response <- function(model, data, predictors, n_grid = 200) {
  med_vals <- apply(data[, predictors, drop = FALSE], 2, median, na.rm = TRUE)
  out_list <- list()
  for (v in predictors) {
    grid_df <- as.data.frame(matrix(rep(med_vals, each = n_grid), nrow = n_grid, byrow = FALSE))
    names(grid_df) <- predictors
    grid_df[[v]] <- seq(min(data[[v]], na.rm = TRUE), max(data[[v]], na.rm = TRUE), length.out = n_grid)
    pred <- predict(model, newdata = grid_df, type = "link", se.fit = TRUE)
    grid_df$fit   <- plogis(pred$fit)
    grid_df$lower <- plogis(pred$fit - 1.96 * pred$se.fit)
    grid_df$upper <- plogis(pred$fit + 1.96 * pred$se.fit)
    grid_df$x <- grid_df[[v]]
    grid_df$variable <- v
    out_list[[v]] <- grid_df[, c("x", "fit", "lower", "upper", "variable")]
  }
  do.call(rbind, out_list)
}

soil_labels <- c("1. Very Low","2. Low","3. Medium-Low","4. Medium-High","5. High","6. Optimal")
soil_colors <- c("#5D3A00","#A66C2D","#D9B38C","#B5E28A","#4CAF50","#1B5E20")

raster_to_df <- function(r, class_labels = soil_labels) {
  df <- as.data.frame(r, xy = TRUE, na.rm = TRUE)
  names(df)[3] <- "class"
  df$class <- factor(df$class, levels = seq_along(class_labels), labels = class_labels)
  df
}

make_spatial_blocks_square <- function(x, raster_layer, size = 2000, k = 5, column = "y") {
  cv_obj <- NULL

  if ("cv_spatial" %in% getNamespaceExports("blockCV")) {
    invisible(utils::capture.output({
      cv_obj <- blockCV::cv_spatial(
        x = x, column = column, r = raster_layer,
        size = size, k = k, selection = "systematic",
        hexagon = FALSE, progress = FALSE, report = FALSE, plot = FALSE
      )
    }))
  } else {
    invisible(utils::capture.output({
      cv_obj <- blockCV::spatialBlock(
        speciesData = x, species = column, rasterLayer = raster_layer,
        theRange = size, k = k, selection = "systematic",
        showBlocks = FALSE
      )
    }))
  }

  if (is.null(cv_obj)) stop("Spatial blocking failed: blockCV did not return an object.")

  k_out <- if (!is.null(cv_obj$k)) cv_obj$k else k
  folds_list <- cv_obj$folds_list

  if (!is.null(cv_obj$folds_ids)) {
    fold_ids <- cv_obj$folds_ids
  } else {
    fold_ids <- rep(NA_integer_, nrow(x))
    for (f in seq_len(k_out)) {
      te <- folds_list[[f]][[2]]
      if (length(te) > 0) fold_ids[te] <- f
    }
  }

  list(
    k = k_out,
    folds_list = folds_list,
    fold_ids = fold_ids,
    blocks = cv_obj$blocks
  )
}

make_block_count_table <- function(block_obj, y) {
  out <- lapply(seq_len(block_obj$k), function(f) {
    tr <- block_obj$folds_list[[f]][[1]]
    te <- block_obj$folds_list[[f]][[2]]
    c(
      train_0 = sum(y[tr] == 0, na.rm = TRUE),
      train_1 = sum(y[tr] == 1, na.rm = TRUE),
      test_0  = sum(y[te] == 0, na.rm = TRUE),
      test_1  = sum(y[te] == 1, na.rm = TRUE)
    )
  })
  out <- as.data.frame(do.call(rbind, out))
  rownames(out) <- seq_len(nrow(out))
  zero_cols <- vapply(out, function(z) all(is.na(z) | z == 0), logical(1))
  out[, !zero_cols, drop = FALSE]
}

get_block_labels <- function(blocks_sf, points = NULL, point_fold_col = NULL) {
  candidate_cols <- c("folds_ids", "folds_id", "foldID", "fold_id", "folds", "fold", "id")
  hit <- candidate_cols[candidate_cols %in% names(blocks_sf)]
  if (length(hit) > 0) return(as.character(blocks_sf[[hit[1]]]))
  if (!is.null(points) && !is.null(point_fold_col) && point_fold_col %in% names(points)) {
    idx <- sf::st_intersects(blocks_sf, points)
    lab <- vapply(idx, function(ii) {
      vals <- points[[point_fold_col]][ii]
      vals <- vals[!is.na(vals)]
      if (!length(vals)) return(NA_character_)
      as.character(names(sort(table(vals), decreasing = TRUE)[1]))
    }, character(1))
    return(lab)
  }
  as.character(seq_len(nrow(blocks_sf)))
}

make_hillshade_df <- function(dem_raster, fact = 4, angle = 45, direction = 315) {
  dem_agg <- terra::aggregate(dem_raster, fact = fact, fun = "mean", na.rm = TRUE)
  slope_hs <- terra::terrain(dem_agg, v = "slope", unit = "radians")
  aspect_hs <- terra::terrain(dem_agg, v = "aspect", unit = "radians")
  hillshade <- terra::shade(slope_hs, aspect_hs, angle = angle, direction = direction)
  hs_df <- as.data.frame(hillshade, xy = TRUE, na.rm = TRUE)
  names(hs_df)[3] <- "hillshade"
  hs_df
}

sample_skewness <- function(x) {
  x <- x[is.finite(x)]
  n <- length(x)
  if (n < 3) return(NA_real_)
  m <- mean(x)
  s <- stats::sd(x)
  if (is.na(s) || s == 0) return(0)
  (n / ((n - 1) * (n - 2))) * sum(((x - m) / s)^3)
}

safe_shapiro_p <- function(x, max_n = 5000, seed = 123) {
  x <- x[is.finite(x)]
  if (length(x) < 3) return(NA_real_)
  if (stats::sd(x) == 0) return(NA_real_)
  if (length(x) > max_n) {
    set.seed(seed)
    x <- sample(x, max_n)
  }
  stats::shapiro.test(x)$p.value
}

format_p_bar <- function(p) {
  ifelse(
    is.na(p), NA_character_,
    ifelse(
      p < 1e-4, "p < 0.0001",
      ifelse(
        p < 0.001, "p < 0.001",
        paste0("p = ", formatC(p, format = "f", digits = 2))
      )
    )
  )
}

format_p_subtitle <- function(p, digits = 3) {
  ifelse(
    is.na(p), NA_character_,
    ifelse(
      p < 1e-4, "< 0.0001",
      ifelse(
        p < 0.001, "< 0.001",
        paste0("= ", formatC(p, format = "f", digits = digits))
      )
    )
  )
}

make_omnibus_subtitle <- function(test_obj, stat_digits = 2, p_digits = 3) {
  if (identical(test_obj$omnibus_label, "Kruskal-Wallis")) {
    paste0(
      "Kruskal-Wallis = ",
      formatC(test_obj$omnibus_statistic, format = "f", digits = stat_digits),
      ", p ",
      format_p_subtitle(test_obj$omnibus_p, digits = p_digits)
    )
  } else if (identical(test_obj$omnibus_label, "ANOVA")) {
    paste0(
      "ANOVA F = ",
      formatC(test_obj$omnibus_statistic, format = "f", digits = stat_digits),
      ", p ",
      format_p_subtitle(test_obj$omnibus_p, digits = p_digits)
    )
  } else {
    paste0(
      test_obj$omnibus_label,
      ": p ",
      format_p_subtitle(test_obj$omnibus_p, digits = p_digits)
    )
  }
}

choose_group_test <- function(df, value_col, group_col, y_step = 0.08) {
  dd <- df %>%
    dplyr::select(all_of(c(value_col, group_col))) %>%
    dplyr::rename(.value = all_of(value_col), .group = all_of(group_col)) %>%
    dplyr::filter(is.finite(.value), !is.na(.group)) %>%
    dplyr::mutate(.group = factor(.group))

  if (nlevels(dd$.group) < 2) stop("At least two groups are required.")
  if (nrow(dd) == 0) stop("No valid rows available for testing.")

  shapiro_tbl <- dd %>%
    dplyr::group_by(.group) %>%
    dplyr::summarise(
      n = dplyr::n(),
      shapiro_p = safe_shapiro_p(.value),
      .groups = "drop"
    )

  levene_p <- tryCatch({
    as.numeric(car::leveneTest(.value ~ .group, data = dd)[1, "Pr(>F)"])
  }, error = function(e) NA_real_)

  normal_ok <- all(!is.na(shapiro_tbl$shapiro_p) & shapiro_tbl$shapiro_p > 0.05)
  variance_ok <- !is.na(levene_p) && levene_p > 0.05

  if (normal_ok && variance_ok) {
    omnibus <- stats::aov(.value ~ .group, data = dd)
    omnibus_p <- summary(omnibus)[[1]][["Pr(>F)"]][1]
    omnibus_statistic <- summary(omnibus)[[1]][["F value"]][1]
    omnibus_label <- "ANOVA"
    method_label <- "Parametric"
    pairwise <- rstatix::pairwise_t_test(
      dd, .value ~ .group,
      p.adjust.method = "holm",
      pool.sd = FALSE
    )
  } else {
    omnibus <- stats::kruskal.test(.value ~ .group, data = dd)
    omnibus_p <- omnibus$p.value
    omnibus_statistic <- unname(omnibus$statistic)
    omnibus_label <- "Kruskal-Wallis"
    method_label <- "Nonparametric"
    pairwise <- rstatix::pairwise_wilcox_test(
      dd, .value ~ .group,
      p.adjust.method = "holm",
      exact = FALSE
    )
  }

  pairwise <- pairwise %>%
    rstatix::add_y_position(
      fun = "max",
      data = dd,
      formula = .value ~ .group,
      step.increase = y_step
    ) %>%
    dplyr::mutate(
      p_label = format_p_bar(p.adj)
    )

  list(
    data = dd,
    shapiro = shapiro_tbl,
    levene_p = levene_p,
    normal_ok = normal_ok,
    variance_ok = variance_ok,
    method_label = method_label,
    omnibus_label = omnibus_label,
    omnibus_statistic = omnibus_statistic,
    omnibus_p = omnibus_p,
    pairwise = pairwise
  )
}

```

# 4. Terrain predictors

```{r terrain}
dem_mask <- terra::mask(terra::crop(dem, terra::vect(riofrio)), terra::vect(riofrio))
predictor_cols <- c("Elevation","Slope","TPI","Wetness_TWI","Dist_Channel_DEM")

predictors_path <- file.path(folder, "PREDICTORS_STACK_V11.tif")

if (file.exists(predictors_path)) {
  predictors <- terra::rast(predictors_path)
  names(predictors) <- predictor_cols
  cat(">>> Predictors LOADED from:", predictors_path, "\n")
} else {
  var_elev <- dem_mask
  names(var_elev) <- "Elevation"
  
  slope <- terra::terrain(dem_mask, v = "slope", unit = "degrees")
  names(slope) <- "Slope"
  
  var_tpi <- terra::terrain(dem_mask, v = "TPI")
  names(var_tpi) <- "TPI"
  
  fdir <- terra::terrain(dem_mask, v = "flowdir")
  fac  <- terra::flowAccumulation(fdir)
  slope_rad <- terra::clamp(slope * pi / 180, lower = 0.001, upper = 1.57)
  var_twi <- log((fac + 1) / tan(slope_rad))
  names(var_twi) <- "Wetness_TWI"
  
  channel_threshold <- 250
  channel_cells <- fac >= channel_threshold
  channel_cells[channel_cells == 0] <- NA
  var_dist_channel <- terra::mask(terra::distance(channel_cells), dem_mask)
  names(var_dist_channel) <- "Dist_Channel_DEM"
  
  predictors <- c(var_elev, slope, var_tpi, var_twi, var_dist_channel)
  terra::writeRaster(predictors, predictors_path, overwrite = TRUE, datatype = "FLT4S")
  cat(">>> Predictors GENERATED and SAVED to:", predictors_path, "\n")
}

cat(">>> PREDICTOR SUMMARY:\n")
print(round(terra::global(predictors, fun = c("mean","sd","min","max"), na.rm = TRUE), 3))
```

# 5. Terrace de-duplication and model dataset

```{r dedup}
terrazas$UTM_East  <- sf::st_coordinates(terrazas)[, 1]
terrazas$UTM_North <- sf::st_coordinates(terrazas)[, 2]
terrazas$coord_key <- make_coord_key(terrazas$UTM_East, terrazas$UTM_North)

tcp <- terrazas %>%
  sf::st_drop_geometry() %>%
  dplyr::group_by(coord_key) %>%
  dplyr::summarise(
    n_periods = dplyr::n_distinct(Period),
    period_label = ifelse(n_periods == 1, dplyr::first(Period), NA_character_),
    .groups = "drop"
  )

cat(">>> DUPLICATE DIAGNOSTIC:\n")
print(
  tcp %>%
    dplyr::count(n_periods, name = "n_coords") %>%
    dplyr::mutate(pct = round(100 * n_coords / sum(n_coords), 1))
)

terrazas_model <- terrazas %>%
  dplyr::group_by(coord_key) %>%
  dplyr::slice(1) %>%
  dplyr::ungroup() %>%
  dplyr::left_join(tcp, by = "coord_key")

terrazas_v <- terra::vect(terrazas_model)

vals_pres <- terra::extract(predictors, terrazas_v, ID = FALSE)
complete_idx <- which(complete.cases(vals_pres))
vals_pres <- vals_pres[complete_idx, ]
terrazas_model <- terrazas_model[complete_idx, ]

vals_pres$y <- 1
vals_pres$UTM_East <- terrazas_model$UTM_East
vals_pres$UTM_North <- terrazas_model$UTM_North
vals_pres$coord_key <- terrazas_model$coord_key
vals_pres$Temporal_Period <- terrazas_model$period_label

cat(" - Terraces with complete predictor values:", nrow(vals_pres), "\n")

n_abs <- nrow(vals_pres) * 2
bg_path <- file.path(folder, "bg_points_V11.rds")

if (file.exists(bg_path)) {
  bg_points <- readRDS(bg_path)
  cat(">>> Background points LOADED from:", bg_path, "\n")
} else {
  set.seed(123)
  bg_points <- terra::spatSample(
    dem_mask, size = n_abs, method = "random",
    na.rm = TRUE, as.points = TRUE, xy = TRUE
  )
  saveRDS(bg_points, bg_path)
  cat(">>> Background points GENERATED and SAVED to:", bg_path, "\n")
}
vals_abs <- terra::extract(predictors, bg_points, ID = TRUE)
vals_abs <- vals_abs[complete.cases(vals_abs[, setdiff(names(vals_abs), "ID")]), ]
coords_bg <- terra::crds(bg_points)[vals_abs$ID, ]

vals_abs$y <- 0
vals_abs$UTM_East <- coords_bg[, 1]
vals_abs$UTM_North <- coords_bg[, 2]
vals_abs$coord_key <- NA_character_
vals_abs$Temporal_Period <- NA_character_
vals_abs$ID <- NULL

data_model <- rbind(vals_pres, vals_abs)
data_scaled <- data_model
data_scaled[, predictor_cols] <- scale(data_model[, predictor_cols])

cat("\n>>> MODEL DATASET:\n")
cat(" - Unique terrace presences:", sum(data_model$y == 1), "\n")
cat(" - Background absences:", sum(data_model$y == 0), "\n")
cat(" - Total:", nrow(data_model), "\n")
```

# 6. GLM specification, VIF diagnostics, and GAM sensitivity

```{r glm-gam}
ff <- y ~ Elevation + Slope + TPI + Wetness_TWI + Dist_Channel_DEM

final_glm_scaled <- glm(ff, data = data_scaled, family = binomial)
final_glm_raw    <- glm(ff, data = data_model,  family = binomial)

vif_final <- car::vif(final_glm_scaled)
glm_prob  <- predict(final_glm_scaled, type = "response")
glm_auc   <- as.numeric(pROC::auc(pROC::roc(data_scaled$y, glm_prob, quiet = TRUE)))

gam_f <- as.formula(
  paste("y ~", paste(paste0("s(", predictor_cols, ", k=5)"), collapse = " + "))
)
model_gam <- mgcv::gam(gam_f, data = data_scaled, family = binomial, method = "REML")
auc_gam <- as.numeric(pROC::auc(pROC::roc(data_scaled$y,
  predict(model_gam, type = "response"), quiet = TRUE)))

cat(">>> VIF DIAGNOSTIC:\n")
print(round(vif_final, 3))
cat(">>> GLM AUC:", round(glm_auc, 3), "| GAM AUC:", round(auc_gam, 3), "\n")
cat(">>> GAM smooth-term summary:\n")
print(summary(model_gam)$s.table)

preferred_auc <- glm_auc
data_scaled$residuals <- residuals(final_glm_scaled, type = "pearson")
data_scaled$fitted    <- fitted(final_glm_scaled)
```

```{r summ-glm}
cat("\n========== NUMERICAL SUMMARY: GLM AND GAM ==========\n")
cat("GLM formula: y ~ Elevation + Slope + TPI + Wetness_TWI + Dist_Channel_DEM\n")
cat("GLM training AUC:", round(glm_auc, 3), "\n")
cat("GAM training AUC:", round(auc_gam, 3), "\n")
cat("AUC improvement (GAM - GLM):", round(auc_gam - glm_auc, 3), "\n")
cat("Max VIF:", round(max(vif_final), 3), "\n")
cat("====================================================\n")
```

# 7. Residual spatial structure

```{r moran}
crs_res <- cbind(data_scaled$UTM_East, data_scaled$UTM_North)
nb_r <- spdep::knn2nb(spdep::knearneigh(crs_res, k = 8))
lw_r <- spdep::nb2listw(nb_r, style = "W")
moran_test <- spdep::moran.test(data_scaled$residuals, lw_r)
set.seed(999)
moran_perm <- spdep::moran.mc(
  data_scaled$residuals, lw_r, nsim = 10000,
  alternative = "two.sided", zero.policy = TRUE
)

resid_sf <- sf::st_as_sf(data_scaled, coords = c("UTM_East","UTM_North"), crs = sf::st_crs(dem))
vgm_res <- gstat::variogram(residuals ~ 1, data = resid_sf, cutoff = 3000, width = 250)
```

# 8. Spatial block partition

```{r spatial-blocks, include=FALSE}
all_pts <- sf::st_as_sf(data_model, coords = c("UTM_East","UTM_North"), crs = sf::st_crs(dem))
all_pts$y <- data_model$y

sb <- make_spatial_blocks_square(all_pts, predictors[[1]], size = 2000, k = 5)
all_pts$spatial_block <- sb$fold_ids

sb_counts <- make_block_count_table(sb, data_model$y)

data_model$spatial_block <- sb$fold_ids
data_scaled$spatial_block <- sb$fold_ids
```

# 9. Spatial cross-validation

```{r spatial-cv}
auc_cv <- rep(NA_real_, sb$k)

for (fold in seq_len(sb$k)) {
  tr <- sb$folds_list[[fold]][[1]]
  te <- sb$folds_list[[fold]][[2]]

  if (length(tr) == 0 || length(te) == 0) next
  if (length(unique(data_model$y[tr])) < 2 || length(unique(data_model$y[te])) < 2) next

  m <- glm(ff, data = data_model[tr, ], family = binomial)
  p <- predict(m, newdata = data_model[te, ], type = "response")
  auc_cv[fold] <- as.numeric(pROC::auc(pROC::roc(data_model$y[te], p, quiet = TRUE)))
}

auc_cv_mean <- mean(auc_cv, na.rm = TRUE)
auc_cv_sd   <- sd(auc_cv, na.rm = TRUE)

cat(">>> SPATIAL CV: Mean AUC =", round(auc_cv_mean, 3), "+-", round(auc_cv_sd, 3), "\n")
cat(">>> Fold AUCs:", paste(round(auc_cv, 3), collapse = ", "), "\n")
```

# 10. Figure 4. GLM standardized coefficients and ROC curve

```{r fig4, fig.width=8, fig.height=12}
cs <- broom::tidy(final_glm_scaled, conf.int = TRUE) %>%
  dplyr::filter(term != "(Intercept)") %>%
  dplyr::mutate(
    Dir = ifelse(estimate > 0, "Positive", "Negative"),
    term = factor(term, levels = term[order(abs(estimate))])
  )

label_map <- c(
  "Elevation" = "Elevation",
  "Slope" = "Slope",
  "TPI" = "TPI",
  "Wetness_TWI" = "TWI",
  "Dist_Channel_DEM" = "Dist. Channel"
)
cs$term_label <- factor(label_map[as.character(cs$term)],
  levels = label_map[as.character(levels(cs$term))])

p4a <- ggplot(cs, aes(x = term_label, y = estimate, fill = Dir)) +
  geom_col(width = 0.6, color = "black", linewidth = 0.4) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2, color = "black") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  scale_fill_manual(values = c(Positive = "#2E86AB", Negative = "#E63946")) +
  coord_flip() +
  labs(
    title = "A) Standardized GLM coefficients",
    x = NULL, y = "Coefficient", fill = "Direction"
  )

roc_obj <- pROC::roc(data_scaled$y, glm_prob, quiet = TRUE)
roc_df <- data.frame(spec = roc_obj$specificities, sens = roc_obj$sensitivities)

p4b <- ggplot(roc_df, aes(x = 1 - spec, y = sens)) +
  geom_line(color = "#2E86AB", linewidth = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "#E63946") +
  labs(
    title = "B) ROC curve",
    subtitle = paste0(
      "Training AUC = ", round(glm_auc, 3),
      " | Spatial CV AUC = ", round(auc_cv_mean, 3)
    ),
    x = "1 - Specificity", y = "Sensitivity"
  )

print(p4a / p4b)
```

```{r summ-fig4}
cat("\n========== NUMERICAL SUMMARY: FIGURE 4 ==========\n")
cat("4A. STANDARDIZED GLM COEFFICIENTS\n")
for (i in 1:nrow(cs)) {
  cat(sprintf(
    "  %s: b=%.3f [%.3f, %.3f] p=%s\n",
    cs$term[i], cs$estimate[i], cs$conf.low[i], cs$conf.high[i], format.pval(cs$p.value[i], 3)
  ))
}
cat("4B. ROC CURVE\n")
cat("   Training AUC:", round(glm_auc, 3), "| Spatial CV AUC:", round(auc_cv_mean, 3), "\n")
cat("=================================================\n")
```

# 11. Figure 5. GLM response plots

```{r fig5, fig.width=16, fig.height=3.5}
rdf <- make_glm_response(final_glm_scaled, data_scaled, predictor_cols)

rdf$variable <- dplyr::recode(rdf$variable,
  "Dist_Channel_DEM" = "Dist. Channel",
  "Wetness_TWI" = "Wetness_TWI"
)

ggplot(rdf, aes(x = x, y = fit)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#B5E28A") +
  geom_line(color = "black", linewidth = 0.8) +
  facet_wrap(~ variable, scales = "free_x", ncol = 5) +
  labs(
    title = "GLM response plots",
    x = "Standardized value", y = "P(terrace)"
  )
```

```{r summ-fig5}
cat("\n========== NUMERICAL SUMMARY: FIGURE 5 ==========\n")
for (v in unique(rdf$variable)) {
  s <- rdf[rdf$variable == v, ]
  cat(sprintf("  %s: P range %.3f to %.3f\n", v, min(s$fit), max(s$fit)))
}
cat("=================================================\n")
```

# 12. Figure 6. Residual variogram and spatial block partition

```{r fig6, fig.width=10, fig.height=14}
p_var <- ggplot(vgm_res, aes(x = dist, y = gamma)) +
  geom_point(color = "#E63946", size = 3) +
  geom_line(color = "#2E86AB", linewidth = 0.7) +
  geom_hline(yintercept = var(data_scaled$residuals), linetype = "dashed", color = "black") +
  labs(
    title = "A) Residual variogram",
    subtitle = paste0(
      "Moran's I = ", round(moran_test$estimate[1], 3),
      ", permutation p ",
      ifelse(moran_perm$p.value < 0.001, "< 0.001", paste0("= ", round(moran_perm$p.value, 3)))
    ),
    x = "Distance (m)", y = "Semivariance"
  )

blocks_sf <- sf::st_as_sf(sb$blocks)
blocks_sf$block_label <- get_block_labels(blocks_sf, all_pts, "spatial_block")
block_centroids <- suppressWarnings(sf::st_centroid(blocks_sf))
block_centroids$block_label <- blocks_sf$block_label

dem_agg <- terra::aggregate(dem_mask, fact = 4, fun = "mean", na.rm = TRUE)
dem_df <- as.data.frame(dem_agg, xy = TRUE, na.rm = TRUE)
names(dem_df)[3] <- "elev"

p_blk <- ggplot() +
  geom_raster(data = dem_df, aes(x = x, y = y, fill = elev)) +
  scale_fill_gradient(low = "gray30", high = "gray95", guide = "none") +
  geom_sf(data = blocks_sf, fill = NA, color = "#E63946", linewidth = 0.6) +
  geom_sf_text(
    data = block_centroids %>% dplyr::filter(!is.na(block_label)),
    aes(label = block_label), size = 5, fontface = "bold", color = "black"
  ) +
  geom_sf(data = riofrio, fill = NA, color = "black", linewidth = 0.8) +
  ggspatial::annotation_scale(
    location = "bl", width_hint = 0.2,
    text_col = "black", line_col = "black", text_cex = 0.8,
    bar_cols = c("black", "white")
  ) +
  ggspatial::annotation_north_arrow(
    location = "tr", which_north = "true",
    style = ggspatial::north_arrow_fancy_orienteering(
      text_col = "black", line_col = "black", fill = c("black", "black")
    )
  ) +
  coord_sf(datum = sf::st_crs(dem)) +
  labs(
    title = "B) Spatial block partition for cross-validation",
    subtitle = paste0(sb$k, " spatial folds, approximately 2 km blocks")
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 13, color = "black"),
    plot.subtitle = element_text(size = 10, color = "black")
  )

print(p_var / p_blk + patchwork::plot_layout(heights = c(1, 1.4)))
```

```{r summ-fig6}
cat("\n========== NUMERICAL SUMMARY: FIGURE 6 ==========\n")
cat("6A. RESIDUAL VARIOGRAM\n")
cat(" - Moran's I:", round(moran_test$estimate[1], 4), "\n")
cat(" - Permutation p (10,000):", format.pval(moran_perm$p.value, 3), "\n")
cat(" - Variogram bins:", nrow(vgm_res), "\n")
cat(" - Semivariance at shortest lag:", round(vgm_res$gamma[1], 4), "\n")
cat(" - Semivariance at longest lag:", round(vgm_res$gamma[nrow(vgm_res)], 4), "\n")
cat("6B. SPATIAL BLOCKS\n")
cat(" - Blocks:", sb$k, "(square, ~2 km)\n")
cat(" - Fold composition:\n")
print(sb_counts)
cat("=================================================\n")
```

# 13. Suitability surface and original quantile classification

```{r suitability}
suit_map <- terra::predict(predictors, final_glm_raw, type = "response")
names(suit_map) <- "Suitability"
vals_map <- terra::values(suit_map, mat = FALSE, na.rm = TRUE)

breaks <- stats::quantile(vals_map, probs = seq(0, 1, length.out = 7), na.rm = TRUE)
breaks[1] <- breaks[1] - 0.001

rcl_mat <- matrix(c(
  breaks[1], breaks[2], 1,
  breaks[2], breaks[3], 2,
  breaks[3], breaks[4], 3,
  breaks[4], breaks[5], 4,
  breaks[5], breaks[6], 5,
  breaks[6], breaks[7], 6
), ncol = 3, byrow = TRUE)

suit_class <- terra::classify(suit_map, rcl_mat, include.lowest = TRUE)
names(suit_class) <- "Suit_Class"

class_freq_raw <- terra::freq(suit_class)
class_freq <- class_freq_raw[class_freq_raw$value %in% 1:6, ]
class_freq$pct <- round(100 * class_freq$count / sum(class_freq$count), 1)

cat(">>> SUITABILITY RANGE:", round(min(vals_map), 3), "to", round(max(vals_map), 3), "\n")
cat(">>> QUANTILE-BASED BREAKS:\n")
print(data.frame(
  Class = 1:6,
  Lower = round(breaks[1:6], 3),
  Upper = round(breaks[2:7], 3),
  Landscape_pct = class_freq$pct
))
```

# 14. Figure 7. Agricultural suitability model and observed vs expected

```{r fig7a, fig.width=10, fig.height=10}
suit_df <- raster_to_df(suit_class)

Figure_7A <- ggplot() +
  geom_raster(data = suit_df, aes(x = x, y = y, fill = class)) +
  scale_fill_manual(values = soil_colors, name = "Agricultural suitability", drop = FALSE) +
  geom_sf(data = rivers, color = "#1E88E5", linewidth = 0.5) +
  geom_sf(data = riofrio, fill = NA, color = "black", linewidth = 0.8) +
  ggspatial::annotation_north_arrow(
    location = "tr", which_north = "true",
    style = ggspatial::north_arrow_fancy_orienteering(
      text_col = "black", line_col = "black", fill = c("black", "black")
    )
  ) +
  ggspatial::annotation_scale(
    location = "bl", bar_cols = c("black", "white"),
    text_col = "black", line_col = "black"
  ) +
  coord_sf(datum = sf::st_crs(dem)) +
  labs(
    title = "A) Agricultural suitability model",
    subtitle = paste0("Spatial CV AUC = ", round(auc_cv_mean, 3))
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 13, color = "black"),
    plot.subtitle = element_text(size = 10, color = "black"),
    legend.text = element_text(color = "black", size = 9),
    legend.title = element_text(color = "black", face = "bold", size = 10)
  )

print(Figure_7A)
```

```{r fig7b, fig.width=10, fig.height=6}
tv <- terra::vect(terrazas_model)
terrazas_model$Suit_Class <- terra::extract(suit_class, tv, ID = FALSE)$Suit_Class

ts <- terrazas_model %>%
  sf::st_drop_geometry() %>%
  dplyr::filter(!is.na(Suit_Class)) %>%
  dplyr::count(Suit_Class, name = "n") %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1))

cmp_landscape <- data.frame(Class = class_freq$value, Pct = class_freq$pct, Type = "Landscape Availability")
cmp_terraces  <- data.frame(Class = ts$Suit_Class,    Pct = ts$pct,        Type = "Terrace Location")
cmp <- rbind(cmp_landscape, cmp_terraces)
cmp$Label <- soil_labels[cmp$Class]
cmp$Label <- factor(cmp$Label, levels = soil_labels)

chisq_terr <- suppressWarnings(
  chisq.test(x = ts$n, p = class_freq$count / sum(class_freq$count))
)

ggplot(cmp, aes(x = Label, y = Pct, fill = Type)) +
  geom_col(position = position_dodge(0.7), width = 0.6, color = "black", linewidth = 0.3) +
  geom_text(
    aes(label = paste0(Pct, "%")),
    position = position_dodge(0.7), vjust = -0.5,
    size = 3, color = "black", fontface = "bold"
  ) +
  scale_fill_manual(values = c("Landscape Availability" = "#BDBDBD", "Terrace Location" = "#E63946")) +
  labs(
    title = "B) Observed vs expected terrace distribution",
    subtitle = paste0(
      "Chi-sq = ", round(chisq_terr$statistic, 2),
      ", df = ", chisq_terr$parameter, ", p ",
      ifelse(chisq_terr$p.value < 0.001, "< 0.001", round(chisq_terr$p.value, 3))
    ),
    x = "Suitability class", y = "%", fill = NULL
  ) +
  theme(legend.position = "bottom")
```

```{r summ-fig7}
cat("\n========== NUMERICAL SUMMARY: FIGURE 7 ==========\n")
cat("7A. AGRICULTURAL SUITABILITY MAP (quantile classes)\n")
for (i in seq_len(nrow(class_freq))) {
  cl <- class_freq$value[i]
  cat(sprintf(
    "   Class %d: %.3f-%.3f, landscape=%.1f%%\n",
    cl, breaks[cl], breaks[cl + 1], class_freq$pct[i]
  ))
}
cat("   Spatial CV AUC:", round(auc_cv_mean, 3), "\n")
cat("7B. CHI-SQUARE (Observed vs Expected)\n")
cat("   Chi-sq:", round(chisq_terr$statistic, 2), "| p:", format.pval(chisq_terr$p.value, 3), "\n")
merged_summ <- merge(
  data.frame(Class = class_freq$value, land_pct = class_freq$pct),
  data.frame(Class = ts$Suit_Class, terr_pct = ts$pct),
  by = "Class", all.x = TRUE
)
merged_summ$terr_pct[is.na(merged_summ$terr_pct)] <- 0
for (i in seq_len(nrow(merged_summ))) {
  cat(sprintf(
    "   Class %d: landscape=%.1f%%, terraces=%.1f%%\n",
    merged_summ$Class[i], merged_summ$land_pct[i], merged_summ$terr_pct[i]
  ))
}
cat("=================================================\n")
```

# 15. Figure 8. Settlement suitability distribution by period

```{r fig8-data}
sp <- sf::st_point_on_surface(sites_sf)
sv <- terra::vect(sp)

ss <- cbind(
  sf::st_drop_geometry(sp),
  terra::extract(suit_class, sv, ID = FALSE)
)
names(ss)[ncol(ss)] <- "Suit_Class"

ss_sum <- ss %>%
  dplyr::filter(!is.na(Suit_Class)) %>%
  dplyr::group_by(Period, Suit_Class) %>%
  dplyr::summarise(n = n(), .groups = "drop") %>%
  dplyr::group_by(Period) %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1))

sw <- ss_sum %>%
  dplyr::select(Period, Suit_Class, n) %>%
  tidyr::pivot_wider(names_from = Suit_Class, values_from = n, values_fill = 0)

chisq_temp <- chisq.test(sw[, -1])
```

```{r fig8, fig.width=16, fig.height=14}
suit_df <- raster_to_df(suit_class)

ss_p <- ss_sum %>%
  dplyr::mutate(
    PS = factor(
      dplyr::case_when(
        grepl("Neguanje", Period) ~ "Neguanje\n(AD 100\u2013700)",
        grepl("Buritaca", Period) ~ "Buritaca\n(AD 700\u20131000)",
        grepl("Tairona",  Period) ~ "Tairona\n(AD 1000\u20131600)"
      ),
      levels = c("Neguanje\n(AD 100\u2013700)", "Buritaca\n(AD 700\u20131000)", "Tairona\n(AD 1000\u20131600)")
    ),
    SC = factor(Suit_Class, levels = 1:6, labels = soil_labels)
  )

p_bars <- ggplot(ss_p, aes(x = SC, y = pct, fill = SC)) +
  geom_col(width = 0.7, color = "black", linewidth = 0.3) +
  geom_text(
    aes(label = paste0(pct, "%")),
    vjust = -0.5,
    size = 3, color = "black", fontface = "bold"
  ) +
  scale_fill_manual(values = soil_colors, guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  facet_wrap(~ PS, ncol = 3) +
  labs(
    title = "Settlement suitability distribution by period",
    subtitle = paste0(
      "Chi\u2013sq = ", round(chisq_temp$statistic, 2),
      ", df = ", chisq_temp$parameter, ", p ",
      ifelse(chisq_temp$p.value < 0.001, "< 0.001", round(chisq_temp$p.value, 3))
    ),
    x = NULL, y = "Sites (%)"
  ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    strip.text = element_text(face = "bold", size = 11)
  )

mk_gg_map <- function(per) {
  pd <- sites_sf %>% dplyr::filter(Period == per)
  short_label <- dplyr::case_when(
    grepl("Neguanje", per) ~ "Neguanje\n(AD 100\u2013700)",
    grepl("Buritaca", per) ~ "Buritaca\n(AD 700\u20131000)",
    grepl("Tairona",  per) ~ "Tairona\n(AD 1000\u20131600)"
  )

  ggplot() +
    geom_raster(data = suit_df, aes(x = x, y = y, fill = class)) +
    scale_fill_manual(values = soil_colors, name = "Suit.", drop = FALSE) +
    geom_sf(data = riofrio, fill = NA, color = "black", linewidth = 0.6) +
    geom_sf(data = pd, fill = "red", color = "blue", linewidth = 0.2) +
    ggspatial::annotation_north_arrow(
      location = "tr", which_north = "true",
      style = ggspatial::north_arrow_fancy_orienteering(
        text_col = "black", line_col = "black", fill = c("black", "black")
      ),
      height = grid::unit(1.2, "cm"), width = grid::unit(1.2, "cm")
    ) +
    ggspatial::annotation_scale(
      location = "bl",
      bar_cols = c("black", "white"), text_col = "black", line_col = "black",
      text_cex = 0.7
    ) +
    coord_sf(datum = sf::st_crs(dem)) +
    labs(title = short_label) +
    theme_void() +
    theme(
      plot.title = element_text(face = "bold", size = 12, color = "black"),
      legend.text = element_text(size = 7),
      legend.title = element_text(size = 8)
    )
}

p_neg <- mk_gg_map("Neguanje (AD 100-700)")
p_bur <- mk_gg_map("Buritaca (AD 700-1000)")
p_tai <- mk_gg_map("Tairona (AD 1000-1600)")

p_maps <- (p_neg + p_bur + p_tai + patchwork::plot_layout(ncol = 3, guides = "collect")) &
  theme(legend.position = "bottom")

print(p_bars / p_maps + patchwork::plot_layout(heights = c(1, 1.2)))
```

```{r summ-fig8}
cat("\n========== NUMERICAL SUMMARY: FIGURE 8 ==========\n")
cat("Temporal chi-sq:", round(chisq_temp$statistic, 2), "| p:", format.pval(chisq_temp$p.value, 3), "\n")
print(as.data.frame(ss_sum))
cat("=================================================\n")
```

# 16. Figure 9. Terrace diameter distributions by suitability zone

```{r fig9-data}
terr_base <- terrazas_model %>%
  sf::st_drop_geometry() %>%
  dplyr::select(-dplyr::any_of(c("Suit_Class", "Suit_Prob")))

terr_suit <- cbind(
  terr_base,
  terra::extract(suit_map, tv, ID = FALSE),
  terra::extract(suit_class, tv, ID = FALSE)
)
colnames(terr_suit)[(ncol(terr_suit) - 1):ncol(terr_suit)] <- c("Suit_Prob", "Suit_Class")

tsz <- terr_suit %>%
  dplyr::filter(!is.na(Suit_Class), !is.na(Diam_Terra)) %>%
  dplyr::mutate(
    zone = factor(
      dplyr::case_when(
        Suit_Class <= 2 ~ "Suboptimal (1-2)",
        Suit_Class <= 4 ~ "Intermediate (3-4)",
        TRUE ~ "Optimal (5-6)"
      ),
      levels = c("Suboptimal (1-2)", "Intermediate (3-4)", "Optimal (5-6)")
    )
  )

global_q90 <- as.numeric(stats::quantile(tsz$Diam_Terra, 0.90, na.rm = TRUE))
global_q95 <- as.numeric(stats::quantile(tsz$Diam_Terra, 0.95, na.rm = TRUE))
global_q99 <- as.numeric(stats::quantile(tsz$Diam_Terra, 0.99, na.rm = TRUE))

shape_summary <- tsz %>%
  dplyr::group_by(zone) %>%
  dplyr::summarise(
    n = dplyr::n(),
    mean = round(mean(Diam_Terra, na.rm = TRUE), 2),
    median = round(median(Diam_Terra, na.rm = TRUE), 2),
    sd = round(sd(Diam_Terra, na.rm = TRUE), 2),
    q75 = round(quantile(Diam_Terra, 0.75, na.rm = TRUE), 2),
    q90 = round(quantile(Diam_Terra, 0.90, na.rm = TRUE), 2),
    q95 = round(quantile(Diam_Terra, 0.95, na.rm = TRUE), 2),
    q99 = round(quantile(Diam_Terra, 0.99, na.rm = TRUE), 2),
    max = round(max(Diam_Terra, na.rm = TRUE), 2),
    skewness = round(sample_skewness(Diam_Terra), 3),
    n_ge_global_q95 = sum(Diam_Terra >= global_q95, na.rm = TRUE),
    pct_ge_global_q95 = round(100 * mean(Diam_Terra >= global_q95, na.rm = TRUE), 2),
    .groups = "drop"
  )

tail_long <- shape_summary %>%
  dplyr::select(zone, q90, q95, q99, max) %>%
  tidyr::pivot_longer(
    cols = c(q90, q95, q99, max),
    names_to = "tail_stat",
    values_to = "diameter"
  ) %>%
  dplyr::mutate(
    tail_stat = factor(
      tail_stat,
      levels = c("q90", "q95", "q99", "max"),
      labels = c("90th percentile", "95th percentile", "99th percentile", "Maximum")
    )
  )
```


```{r fig9-tests, include=FALSE}
fig9_test <- choose_group_test(tsz, "Diam_Terra", "zone", y_step = 0.10)
```

```{r fig9, fig.width=14, fig.height=11}
zone_cols <- c(
  "Suboptimal (1-2)" = "#E63946",
  "Intermediate (3-4)" = "#F4A261",
  "Optimal (5-6)" = "#2A9D8F"
)

p9a <- ggplot(fig9_test$data, aes(x = .group, y = .value, fill = .group)) +
  geom_boxplot(
    width = 0.22, color = "black", linewidth = 0.35,
    outlier.shape = 16, outlier.size = 1.2, outlier.alpha = 0.8
  ) +
  ggpubr::stat_pvalue_manual(
    fig9_test$pairwise,
    label = "p_label",
    tip.length = 0.012,
    bracket.size = 0.4,
    bracket.shorten = 0.03,
    size = 4,
    vjust = 0.2
  ) +
  scale_fill_manual(values = zone_cols) +
  scale_y_continuous(expand = expansion(mult = c(0.04, 0.20))) +
  labs(
    title = "A) Boxplot by suitability zone",
    subtitle = make_omnibus_subtitle(fig9_test),
    x = "Zone", y = "Terrace diameter (m)"
  ) +
  theme(legend.position = "none")

p9b <- ggplot(tsz, aes(x = Diam_Terra, fill = zone)) +
  geom_histogram(bins = 35, color = "white", alpha = 0.9) +
  facet_wrap(~ zone, ncol = 1, scales = "free_y") +
  scale_fill_manual(values = zone_cols) +
  labs(
    title = "B) Histograms by suitability zone",
    subtitle = "Right-skew and upper-tail structure",
    x = "Terrace diameter (m)", y = "Frequency"
  ) +
  theme(legend.position = "none")

p9c <- ggplot(tsz, aes(x = Diam_Terra, color = zone)) +
  stat_ecdf(linewidth = 1) +
  geom_vline(xintercept = global_q95, linetype = "dashed", color = "black") +
  scale_color_manual(values = zone_cols) +
  labs(
    title = "C) ECDF by suitability zone",
    subtitle = paste0("Dashed line = global 95th percentile (", round(global_q95, 2), " m)"),
    x = "Terrace diameter (m)", y = "Cumulative proportion", color = "Zone"
  )

p9d <- ggplot(tail_long, aes(x = tail_stat, y = diameter, group = zone, color = zone)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 3) +
  scale_color_manual(values = zone_cols) +
  labs(
    title = "D) Upper-tail comparison",
    subtitle = "90th, 95th, 99th percentiles and maximum values",
    x = NULL, y = "Terrace diameter (m)", color = "Zone"
  )

print((p9a + p9b) / (p9c + p9d))
```


```{r summ-fig9}
cat("
========== NUMERICAL SUMMARY: FIGURE 9 ==========
")
cat("Global thresholds:
")
cat(" - 90th percentile:", round(global_q90, 2), "m
")
cat(" - 95th percentile:", round(global_q95, 2), "m
")
cat(" - 99th percentile:", round(global_q99, 2), "m

")

cat("Automatic assumption screening:
")
print(fig9_test$shapiro)
cat("Levene p:", format.pval(fig9_test$levene_p, 3), "
")
cat("Selected framework:", fig9_test$method_label, "
")
cat("Omnibus test:", fig9_test$omnibus_label, "
")
cat("Omnibus p:", format.pval(fig9_test$omnibus_p, 3), "

")

cat("Pairwise comparisons (Holm-adjusted):
")
print(fig9_test$pairwise %>% dplyr::select(group1, group2, p, p.adj, p_label, y.position))
cat("
Distribution summary:
")
print(shape_summary)

largest_terraces <- tsz %>%
  dplyr::arrange(dplyr::desc(Diam_Terra)) %>%
  dplyr::select(zone, Diam_Terra, Suit_Class, Suit_Prob) %>%
  dplyr::slice_head(n = 15)

cat("
Top 15 largest terraces:
")
print(largest_terraces)
cat("=================================================
")
```

# 17. Figure 10. Optimal land access by period

```{r catchment, include=FALSE}
oz <- suit_class
oz[oz < 5] <- NA

sps <- suppressWarnings(sf::st_point_on_surface(sites_sf))
sps$SID <- seq_len(nrow(sps))

survey_union <- sf::st_union(sf::st_make_valid(riofrio))

site_buffers <- suppressWarnings(
  sf::st_intersection(
    sf::st_buffer(sps, 1000),
    survey_union
  )
)

site_buffers <- site_buffers %>%
  dplyr::group_by(SID, Period) %>%
  dplyr::summarise(geometry = sf::st_union(geometry), .groups = "drop")

bv <- terra::vect(site_buffers)

cell_area_ha <- prod(terra::res(oz)) / 10000
opt_cells <- terra::extract(oz, bv, fun = function(x, ...) sum(!is.na(x)), ID = TRUE)
opt_value_col <- setdiff(names(opt_cells), "ID")[1]
opt_cells$SID <- site_buffers$SID[opt_cells$ID]

buffer_area <- site_buffers %>%
  dplyr::mutate(TB_ha = as.numeric(sf::st_area(.) / 10000)) %>%
  sf::st_drop_geometry() %>%
  dplyr::select(SID, TB_ha)

buffer_optimal <- opt_cells %>%
  dplyr::transmute(SID, OC = .data[[opt_value_col]]) %>%
  dplyr::group_by(SID) %>%
  dplyr::summarise(OC = sum(OC, na.rm = TRUE), .groups = "drop")

soa <- sps %>%
  sf::st_drop_geometry() %>%
  dplyr::select(SID, Period) %>%
  dplyr::left_join(buffer_area, by = "SID") %>%
  dplyr::left_join(buffer_optimal, by = "SID") %>%
  dplyr::mutate(
    OC = dplyr::coalesce(OC, 0),
    OA_ha = OC * cell_area_ha,
    Pct = ifelse(!is.na(TB_ha) & TB_ha > 0, 100 * OA_ha / TB_ha, NA_real_)
  )

soa_valid <- soa %>%
  dplyr::filter(is.finite(Pct), !is.na(Period)) %>%
  dplyr::mutate(
    PS = period_factor(period_short(Period))
  )

sps_coords <- sf::st_coordinates(sps)
soa_valid$X <- sps_coords[match(soa_valid$SID, sps$SID), 1]
soa_valid$Y <- sps_coords[match(soa_valid$SID, sps$SID), 2]

soa_sf <- sf::st_as_sf(soa_valid, coords = c("X", "Y"), crs = sf::st_crs(dem), remove = FALSE)
soa_sf$y <- 1

tryCatch({
  site_blocks <- make_spatial_blocks_square(soa_sf, predictors[[1]],
    size = 2000, k = min(5, nrow(soa_sf)), column = "y")
  soa_valid$spatial_block <- site_blocks$fold_ids
}, error = function(e) {
  set.seed(999)
  soa_valid$spatial_block <<- sample(1:5, nrow(soa_valid), replace = TRUE)
})

block_bootstrap_medians <- function(df, nb = 10000, seed = 321) {
  set.seed(seed)
  bids <- unique(stats::na.omit(df$spatial_block))
  pids <- sort(unique(as.character(df$PS)))
  out <- matrix(NA_real_, nrow = nb, ncol = length(pids))
  colnames(out) <- pids
  for (b in seq_len(nb)) {
    si <- sample(bids, length(bids), replace = TRUE)
    sd <- dplyr::bind_rows(lapply(si, function(id) df[df$spatial_block == id, , drop = FALSE]))
    m <- tapply(sd$Pct, as.character(sd$PS), stats::median, na.rm = TRUE)
    out[b, names(m)] <- m
  }
  out
}

bm <- block_bootstrap_medians(soa_valid)
bs <- data.frame(
  PS = colnames(bm),
  boot_median = apply(bm, 2, stats::median, na.rm = TRUE),
  ci_low  = apply(bm, 2, stats::quantile, probs = 0.025, na.rm = TRUE),
  ci_high = apply(bm, 2, stats::quantile, probs = 0.975, na.rm = TRUE)
) %>%
  dplyr::mutate(PS = factor(PS, levels = levels(soa_valid$PS)))
```

```{r fig10-tests, include=FALSE}
fig10_test <- choose_group_test(soa_valid, "Pct", "PS", y_step = 0.09)
```

```{r fig10, fig.width=12, fig.height=6}
boot_colors <- c(
  "Neguanje\n(AD 100-700)" = "#2E86AB",
  "Buritaca\n(AD 700-1000)" = "#F4A261",
  "Tairona\n(AD 1000-1600)" = "#E63946"
)

p10a <- ggplot(fig10_test$data, aes(x = .group, y = .value, fill = .group)) +
  geom_boxplot(
    width = 0.5, color = "black", linewidth = 0.35,
    outlier.shape = 16, outlier.size = 1, outlier.alpha = 0.6
  ) +
  ggpubr::stat_pvalue_manual(
    fig10_test$pairwise,
    label = "p_label",
    tip.length = 0.012,
    bracket.size = 0.4,
    bracket.shorten = 0.03,
    size = 3.5,
    vjust = 0.2
  ) +
  scale_fill_manual(values = boot_colors) +
  scale_y_continuous(expand = expansion(mult = c(0.04, 0.18))) +
  labs(
    title = "A) Optimal land access by period",
    subtitle = make_omnibus_subtitle(fig10_test),
    x = NULL, y = "Optimal land in 1 km catchment (%)"
  ) +
  theme(legend.position = "none")

p10b <- ggplot(bs, aes(x = PS, y = boot_median, color = PS)) +
  geom_hline(yintercept = median(soa_valid$Pct, na.rm = TRUE),
    linetype = "dashed", color = "black", linewidth = 0.4) +
  geom_pointrange(aes(ymin = ci_low, ymax = ci_high),
    linewidth = 1, size = 1.2) +
  geom_text(aes(label = paste0(round(boot_median, 1), "%")),
    vjust = -1.5, size = 4, color = "black", fontface = "bold") +
  geom_text(aes(label = paste0("[", round(ci_low, 1), " - ", round(ci_high, 1), "]")),
    vjust = 2.5, size = 3, color = "black") +
  scale_color_manual(values = boot_colors) +
  scale_y_continuous(limits = c(
    min(bs$ci_low) - 5,
    max(bs$ci_high) + 8)) +
  labs(
    title = "B) Spatial bootstrap",
    subtitle = "Medians with 95% CIs (10,000 iterations)",
    x = NULL, y = "Median % optimal land"
  ) +
  theme(legend.position = "none")

print(p10a + p10b)
```

```{r summ-fig10}
cat("
========== NUMERICAL SUMMARY: FIGURE 10 ==========
")
cat("Automatic assumption screening:
")
print(fig10_test$shapiro)
cat("Levene p:", format.pval(fig10_test$levene_p, 3), "
")
cat("Selected framework:", fig10_test$method_label, "
")
cat("Omnibus test:", fig10_test$omnibus_label, "
")
cat("Omnibus p:", format.pval(fig10_test$omnibus_p, 3), "

")
cat("Pairwise comparisons (Holm-adjusted):
")
print(fig10_test$pairwise %>% dplyr::select(group1, group2, p, p.adj, p_label, y.position))
cat("
Spatially blocked bootstrap medians (10,000 iterations):
")
print(bs %>% dplyr::select(PS, boot_median, ci_low, ci_high) %>%
  dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 2))))
cat("=================================================
")
```

# 18. Summary tables

```{r table1, results='asis'}
gam_st <- as.data.frame(summary(model_gam)$s.table)
gam_st$Variable <- gsub("s\\((.*)\\)", "\\1", rownames(gam_st))

table1 <- cs %>%
  dplyr::transmute(
    Variable = as.character(term),
    VIF = round(vif_final[as.character(term)], 2),
    GLM_beta = round(estimate, 3),
    GLM_95CI = paste0("[", round(conf.low, 3), ", ", round(conf.high, 3), "]"),
    GLM_p = format.pval(p.value, digits = 2),
    GAM_edf = round(gam_st$edf[match(as.character(term), gam_st$Variable)], 2),
    GAM_p = format.pval(gam_st$`p-value`[match(as.character(term), gam_st$Variable)], digits = 2)
  )

cat("\n========== TABLE 1: PREDICTOR DIAGNOSTICS ==========\n")
cat("GLM training AUC:", round(glm_auc, 3),
    "| GAM AUC:", round(auc_gam, 3),
    "| Spatial CV AUC:", round(auc_cv_mean, 3), "+-", round(auc_cv_sd, 3), "\n\n")
print(knitr::kable(table1, format = "simple", align = "lcccccc",
  caption = "Table 1. Predictor diagnostics: VIF, GLM coefficients, and GAM smooth terms"))
cat("\n=====================================================\n")
```

```{r table2, results='asis'}
period_levels <- c("Neguanje (AD 100-700)", "Buritaca (AD 700-1000)", "Tairona (AD 1000-1600)")
period_short_labels <- c("Neguanje", "Buritaca", "Tairona")

site_n <- sites_sf %>%
  sf::st_drop_geometry() %>%
  dplyr::count(Period, name = "N_sites")

terr_period <- terrazas %>%
  sf::st_drop_geometry() %>%
  dplyr::filter(!is.na(Diam_Terra), !is.na(Period)) %>%
  dplyr::group_by(Period) %>%
  dplyr::summarise(
    N_terraces = dplyr::n(),
    Diam_median = round(median(Diam_Terra, na.rm = TRUE), 1),
    Diam_skew = round(sample_skewness(Diam_Terra), 2),
    .groups = "drop"
  )

boot_tbl <- bs %>%
  dplyr::mutate(
    Period = dplyr::case_when(
      grepl("Neguanje", PS) ~ "Neguanje (AD 100-700)",
      grepl("Buritaca", PS) ~ "Buritaca (AD 700-1000)",
      grepl("Tairona",  PS) ~ "Tairona (AD 1000-1600)"
    ),
    Boot_median = round(boot_median, 1),
    Boot_CI = paste0("[", round(ci_low, 1), ", ", round(ci_high, 1), "]")
  ) %>%
  dplyr::select(Period, Boot_median, Boot_CI)

table2 <- site_n %>%
  dplyr::left_join(terr_period, by = "Period") %>%
  dplyr::left_join(boot_tbl, by = "Period") %>%
  dplyr::mutate(
    Period = factor(Period, levels = period_levels, labels = period_short_labels)
  ) %>%
  dplyr::arrange(Period) %>%
  dplyr::select(Period, N_sites, N_terraces, Diam_median, Diam_skew, Boot_median, Boot_CI)

cat("\n========== TABLE 2: PERIOD-LEVEL SUMMARY ==========\n")
cat("Temporal chi-sq:", round(chisq_temp$statistic, 2),
    "| p:", format.pval(chisq_temp$p.value, 3), "\n\n")
print(knitr::kable(table2, format = "simple", align = "lcccccc",
  col.names = c("Period", "Sites", "Terraces", "Diam. med.", "Diam. skew",
                "Boot. med. (%)", "Boot. 95% CI"),
  caption = "Table 2. Period-level temporal stability and distributional shape"))
cat("\n=====================================================\n")
```

# 19. Final numerical synthesis

```{r final-summary}
cat("=================================================
")
cat("FINAL SYNTHESIS - JAA_R_NOTEBOOK_V12
")
cat("=================================================
")
cat("Model: GLM with DEM-derived channel distance
")
cat("Training AUC:", round(glm_auc, 3), "
")
cat("GAM AUC:", round(auc_gam, 3), "| GAM-GLM difference:", round(auc_gam - glm_auc, 3), "
")
cat("Spatial CV AUC:", round(auc_cv_mean, 3), "+-", round(auc_cv_sd, 3), "
")
cat("Max VIF:", round(max(vif_final), 3), "
")
cat("Moran's I (residuals):", round(moran_test$estimate[1], 4),
    "| permutation p:", format.pval(moran_perm$p.value, 3), "
")
cat("Terraces vs landscape chi-square:", round(chisq_terr$statistic, 2),
    "| p:", format.pval(chisq_terr$p.value, 3), "
")
cat("Temporal site distribution chi-square:", round(chisq_temp$statistic, 2),
    "| p:", format.pval(chisq_temp$p.value, 3), "
")
cat("Figure 9 omnibus:", fig9_test$omnibus_label,
    "| p:", format.pval(fig9_test$omnibus_p, 3), "
")
cat("Figure 10 omnibus:", fig10_test$omnibus_label,
    "| p:", format.pval(fig10_test$omnibus_p, 3), "
")
cat("Bootstrap medians (10,000 spatially blocked iterations):
")
print(bs %>% dplyr::select(PS, boot_median, ci_low, ci_high) %>%
  dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 2))))
cat("Upper-tail terrace summary:
")
print(shape_summary %>% dplyr::select(zone, median, q90, q95, q99, max, skewness, pct_ge_global_q95))
cat("=================================================
")
```
